Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-02-24 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex
Added a note on some pitfalls to avoid when thinking about Sobolev spaces.

\textbf{(3) Why we study.} Laplacian eigenmaps is thus an example of a \emph{spectral series} estimator. Series estimators are one of the most classical ways to perform non-parametric regression. Traditionally, series estimators are defined with respect to a fixed set of basis functions, orthogonalized with respect to some pre-determined reference measure $Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the distribution $P$ in a rich manner.

The study of (classical) spectral series estimators is rich, and their theoretical optimality is well-understood \textcolor{red}{(references)}. However, in practice these estimators suffer from some serious drawbacks.

@@

Broadly speaking, we show that when the regression function $f_0$ is smooth, in the sense of having bounded derivatives in the Sobolev sense, Laplacian eigenmaps is statistically minimax optimal. This statement holds for different relations between the dimension $d$ and number of derivatives $s$, depending on the problem (estimation) \textcolor{red}{(TODO): Add two tables summarizing your major results here.}

@@

@@

\textbf{(1) Sobolev norms, semi-norms, and balls.} 
The Sobolev class $H^s(P)$ consists of all those functions $f \in \Leb^2(P)$ which have weak partial derivatives of all orders $j = 1,\ldots,s$, all of which are bounded in $\Leb^2(P)$ norm. The Sobolev norm $\|f\|_{H^s(P)}$ is defined according to
\begin{equation*}
\|f\|_{H^s(P)}^2 := \|f\|_{\Leb^2(P)}^2 + \sum_{j = 1}^{s} \|D^jf\|_{\Leb^2(P)}^2,
\end{equation*}
where $D^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$, and we use the multiindex notation $D^{\alpha}f = \partial^j f/(\partial^{\alpha_1}x^1\ldots \partial^{\alpha_d}x^d)$ for multiindex $\alpha$. The Sobolev ball of radius $M$ consists of all those functions $f \in H^s(P)$ with Sobolev norm no greater than $M$,
\begin{equation*}
H^s(\mc{X};M) := \{f \in H^s(P): \|f\|_{H^s(P)} \leq M\}
\end{equation*}

@@

\subsection{First-order Sobolev classes}
\label{sec:first_order_sobolev_classes}
To begin, we handle the case where $f_0 \in H^1(\mc{X})$.

\textbf{(1) Laplacian eigenmaps is an optimal estimator.} When the graph radius $\varepsilon$ and number of eigenvectors $K$ are chosen appropriately, the estimator $\wh{f}$ has (in-sample) optimal risk (up to constant factors) $\|\wh{f} - f_0\|_n^2 \lesssim n^{-2/(2 + d)}$. 
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{3}
	\item 
	\label{asmp:estimation_fo} 
	The graph radius $\varepsilon$ and the number of eigenvectors $K$ satisfy the following inequalities:
	\begin{equation}\\
	\label{eqn:radius_estimation_fo} 
	C_1 \biggl(\frac{\log n}{n}\biggr)^{1/d} \leq \varepsilon \leq \min\{i_0,K^{-1/d}\},
	\end{equation}
	and 
	\begin{equation}
	\label{eqn:eigenvector_estimation_fo} 
	c_2 (M^2 n)^{2d/(2 + d)}\leq K \leq C_2 (M^2 n)^{2d/(2 + d)}.
	\end{equation}
\end{enumerate}
\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_fo}
	Suppose the general assumptions~\ref{asmp:domain}-\ref{asmp:sobolev_radius} are satisfied. Then for any $f_0 \in H^1(P,M)$, if Laplacian eigenmaps is run with parameters $\varepsilon$ and $K$ satisfying~\ref{asmp:estimation_fo}, with probability at least $1 - \delta - \textcolor{red}{(???)}$ it follows that
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_fo}
	\|\wh{f} - f_0\|_n^2 \leq \frac{C}{\delta}M^{2d/(2 + d)}n^{-2/(2 + d)}
	\end{equation}
\end{theorem}
Some remarks:
\begin{itemize}
	\item \textbf{(1.1) How to tune hyper-parameters.} We note that the ranges~\eqref{eqn:radius_estimation_fo} and~\eqref{eqn:eigenvector_estimation_fo} depend on unknown quantities such as $f_0$ and $P$. In practice, one typically tunes hyper-parameters by sample-splitting or cross-validation. However, because the estimator $\wh{f}$ is defined only in-sample, such approaches cannot be straightforwardly applied to select the graph radius $\varepsilon$, or number of eigenvectors $K$. We discuss out-of-sample extensions of $\wh{f}$ later, in Section~\ref{sec:out_of_sample}.
	\item \textbf{(1.2) The dependence on failure probability can be improved.} The upper bound on in-sample risk given by Equation~\eqref{eqn:laplacian_eigenmaps_estimation_fo} depends on the pre-factor $C/\delta$, and holds with probability $1 - \delta - \textcolor{red}{(?)}$. When $f_0 \in C^1(\mc{X};M)$, we can improve these dependencies, changing the pre-factor in~\eqref{eqn:laplacian_eigenmaps_estimation_fo} to $C(1 + 1/\delta)$, and showing that the bound holds with probability at least $1 - C\delta^2/n$. 
\end{itemize}

\textbf{(2) Laplacian eigenmaps is an optimal test.} Consider the test $\varphi = \1\{\wh{T} \geq t_{\delta}\}$, where $t_{\delta}$ is the threshold
\begin{equation*}
t_{\delta} := \frac{K}{n} + \frac{2K}{\sqrt{n}}.
\end{equation*}
The choice of $t_{\delta}$ guarantees that $\varphi$ is a level-$\delta$ test. Under appropriate choices of $\varepsilon$ and $K$, the test $\varphi$ has non-negligible power (up to constant factors) against alternatives separated the null by at least $\|f_0\|_{P}^2 \gtrsim M^{2d/(4 + d)}n^{-4/(4 + d)}$.

\begin{theorem}
	\label{thm:laplacian_eigenmaps_testing_fo}
	\textcolor{red}{(TODO)}
\end{theorem}
Some remarks:
\begin{itemize}
	\item \textbf{(2.1) When $d \geq 4$, the testing problem is different.} When $d \geq 4$, it is not the case that $\Leb^4(\mc{X})$ continuously embeds into $H^1(\mc{X})$, and testing rates over $H^1(\mc{X})$ are not known. On the other hand, if we explicitly assume $f_0 \in \Leb^4(\mc{X})$, then simply taking $T = \|Y\|_n^2$ achieves the optimal rates.
	\item \textbf{(2.2) Both of the remarks after Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} also apply to testing.}
\end{itemize}

>>>>

\textbf{(3) Why we study.} Laplacian eigenmaps is thus an example of a \emph{spectral series} estimator. A spectral series estimator is a special case of an orthogonal series estimator, one of the most classical methods of non-parametric regression. Traditionally, orthogonal series estimators are defined with respect to a fixed set of basis functions, orthogonalized with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series estimators are by this point well-understood \textcolor{red}{(references)}. In particular, they are minimax optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, in practice these estimators suffer from some serious drawbacks. Finding the eigenfunctions of $\Delta_Q$ is in general non-trivial. The statistical optimality of such estimators also hinges on the reference measure $Q$ being equal to $P$, an unrealistic assumption when $P$ is unknown. If $Q \neq P$, the standard recommendation is to instead do least squares---this can be statistically rate-optimal, but is numerically unstable when dimension $d$ is even moderately large.

@@

Broadly speaking, we show that when the regression function $f_0$ is smooth, in the sense of having bounded derivatives in the Sobolev sense, Laplacian eigenmaps is statistically minimax optimal. This statement holds for different relations between the dimension $d$ and number of derivatives $s$, depending on the problem (estimation or testing). \textcolor{red}{(TODO): Add two tables summarizing your major results here.}

@@

\textcolor{red}{(TODO)}

@@

\textbf{(1) Sobolev norms, semi-norms, and balls.} 
The Sobolev class $H^s(P)$ consists of all those functions $f \in \Leb^2(P)$ which have weak partial derivatives of all orders $j = 1,\ldots,s$, all of which are bounded in $\Leb^2(P)$ norm. The Sobolev norm $\|f\|_{H^s(P)}$ is defined according to
\begin{equation*}
\|f\|_{H^s(P)}^2 := \|f\|_{\Leb^2(P)}^2 + \sum_{j = 1}^{s} \|D^jf\|_{\Leb^2(P)}^2,
\end{equation*}
where $D^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$, and we use the multiindex notation $D^{\alpha}f = \partial^j f/(\partial^{\alpha_1}x^1\ldots \partial^{\alpha_d}x^d)$ for multiindex $\alpha$. The Sobolev ball of radius $M$ consists of all those functions $f \in H^s(P)$ with Sobolev norm no greater than $M$,
\begin{equation*}
H^s(P;M) := \{f \in H^s(P): \|f\|_{H^s(P)} \leq M\}
\end{equation*}

@@

\subsection{First-order Sobolev classes}
\label{sec:first_order_sobolev_classes}
To begin, we handle the case where $f_0 \in H^1(P)$. We show that $\wh{f}$ and a test based on $\wh{T}$ are minimax optimal, for all values of $d$, and under no additional assumptions on the data generating process, i.e. on $P$ or $f_0$.

\paragraph{Estimation.} When the graph radius $\varepsilon$ and number of eigenvectors $K$ are chosen appropriately, the estimator $\wh{f}$ has (in-sample) optimal risk (up to constant factors) $\|\wh{f} - f_0\|_n^2 \lesssim n^{-2/(2 + d)}$. 
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{3}
	\item 
	\label{asmp:parameters_estimation_fo} 
	The graph radius $\varepsilon$ and the number of eigenvectors $K$ satisfy the following inequalities:
	\begin{equation}\\
	\label{eqn:radius_fo} 
	C_0 \biggl(\frac{\log n}{n}\biggr)^{1/d} \leq \varepsilon \leq \min\{i_0,K^{-1/d}\},
	\end{equation}
	and 
	\begin{equation}
	\label{eqn:eigenvector_estimation_fo} 
	c_1 (M^2 n)^{2d/(2 + d)}\leq K \leq C_1 (M^2 n)^{2d/(2 + d)}.
	\end{equation}
\end{enumerate}
\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation_fo}
	Suppose $f_0 \in H^1(P,M)$. If the Laplacian eigenmaps estimator $\wh{f}$ is computed with parameters $\varepsilon$ and $K$ that satisfy~\ref{asmp:parameters_estimation_fo}, then the following statement holds for any $\delta \in (0,1)$: with probability at least $1 - \delta - \textcolor{red}{(???)}$,
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_estimation_fo}
	\|\wh{f} - f_0\|_n^2 \leq \frac{C}{\delta}M^{2d/(2 + d)}n^{-2/(2 + d)}
	\end{equation}
\end{theorem}
Some remarks:
\begin{itemize}
	\item \textbf{(1.1) How to tune hyper-parameters.} We note that the ranges~\eqref{eqn:radius_fo} and~\eqref{eqn:eigenvector_estimation_fo} depend on unknown quantities such as $f_0$ and $P$. In practice, one typically tunes hyper-parameters by sample-splitting or cross-validation. However, because the estimator $\wh{f}$ is defined only in-sample, such approaches cannot be straightforwardly applied to select the graph radius $\varepsilon$, or number of eigenvectors $K$. We discuss out-of-sample extensions of $\wh{f}$ later, in Section~\ref{sec:out_of_sample}.
	\item \textbf{(1.2) The dependence on failure probability can be improved.} The upper bound on in-sample risk given by Equation~\eqref{eqn:laplacian_eigenmaps_estimation_fo} depends on the pre-factor $C/\delta$, and holds with probability $1 - \delta - \textcolor{red}{(?)}$. When $f_0 \in C^1(\mc{X};M)$, we can improve these dependencies, changing the pre-factor in~\eqref{eqn:laplacian_eigenmaps_estimation_fo} to $C(1 + 1/\delta)$, and showing that the bound holds with probability at least $1 - C\delta^2/n$. 
\end{itemize}

\textbf{(2) Laplacian eigenmaps is an optimal test.} Consider the test $\varphi = \1\{\wh{T} \geq t_{\delta}\}$, where $t_{\delta}$ is the threshold
\begin{equation*}
t_{\delta} := \frac{K}{n} + \frac{2K}{\sqrt{n}}.
\end{equation*}
The choice of $t_{\delta}$ guarantees that $\varphi$ is a level-$\delta$ test. Under appropriate choices of $\varepsilon$ and $K$, the test $\varphi$ has non-negligible power (up to constant factors) against alternatives separated the null by at least $\|f_0\|_{P}^2 \gtrsim M^{2d/(4 + d)}n^{-4/(4 + d)}$, whenever $d < 4$.

\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{4}
	\item 
	\label{asmp:parameters_testing_fo}
	The graph radius $\varepsilon$ and the number of eigenvectors $K$ satisfy~\eqref{eqn:radius_fo}. Additionally,
	\begin{equation}
	\label{eqn:eigenvector_testing_fo}
	c_2 (M^2 n)^{2d/(4 + d)}\leq K \leq C_2 (M^2 n)^{2d/(4 + d)}.
	\end{equation}
\end{enumerate}
Let $p(d) = \min\{1,2/d\}$.
\begin{theorem}
	\label{thm:laplacian_eigenmaps_testing_fo}
	Suppose $f_0 \in H^1(\mc{X};M)$. If the Laplacian eigenmaps test $\varphi$ is computed with parameters $\varepsilon$ and $K$ that satisfy~\ref{asmp:parameters_testing_fo}, then the following statement holds for any $\delta \in (0,1)$: if $f_0$ satisfies
	\begin{equation}
	\label{eqn:laplacian_eigenmaps_testing_criticalradius_fo}
	\|f_0\|_P^2 \geq \frac{C}{\delta} \max\biggl\{\frac{1}{\delta}M^{2d/(4 + d)}n^{-4/(4 + d)}, M n^{-p(d)}\biggr\},
	\end{equation}
	then $\varphi$ has Type II error upper bounded by 
	\begin{equation}
	\Pbb_{f_0}(\varphi = 0) \leq \delta + \textcolor{red}{(...)}
	\end{equation}
\end{theorem}
Some remarks:
\begin{itemize}
	\item \textbf{(2.1) When $d \geq 4$, the testing problem is different.} When $d \geq 4$, it is not the case that $\Leb^4(\mc{X})$ continuously embeds into $H^1(\mc{X})$, and testing rates over $H^1(\mc{X})$ are not known. On the other hand, if we explicitly assume $f_0 \in \Leb^4(\mc{X})$, then simply taking $T = \|Y\|_n^2$ achieves the optimal rates.
	\item \textbf{(2.2) Both of the remarks after Theorem~\ref{thm:laplacian_eigenmaps_estimation_fo} also apply to testing.}
\end{itemize}
